On Optimality of Finite Element Integration

### 2016-01-23 TOMS_Paper/assembly.tex
Other improvements to sharing elimination description

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. Often, the optimal implementation strategy for these operations is to be determined out of two alternatives. To illustrate the concept, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. The tensor operation will reduce to $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on loop $L_i$. To minimize the operation count for sub-expressions of this kind, we have two options.
\begin{Strategy}
\label{strategy:i}
Eliminating temporal sharing through generalized code motion.
\end{Strategy}
\begin{Strategy}
\label{strategy:ii}
Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
\end{Strategy}
In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count, since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed. On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$. In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion. Note that a second example of application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}. These examples motivate the introduction of a particular class of expressions.
\begin{Def}[Structured expression]
\label{def:struct-expr}
An expression is \textit{structured} with respect to a set of loops $L$ if and only if there is no ambiguity in spatial sharing elimination for the symbols depending on $L$.
\end{Def}
The fundamental observation is that if a loop nest $L$ is multilinear, then all of its expressions will be structured with respect to $L$. In finite element integration loop nests, for example, a factorization of test and trial functions will always result in exposing temporal sharing (Strategy~\ref{strategy:ii}). This property will play a key role in Algorithm~\ref{algo:sharing-elimination}, which targets quasi-optimality. Finally, we observe that it is remarkable that structured sub-expressions sometimes characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. It makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\end{itemize}

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree to factorize identical terms within additions. \\ 
\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.

@@

\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i} to the multilinear operands. \\
\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied. This is a consequence of the expression being structured with respect to $[L_j, L_k]$.}

@@

\begin{Rem*}
\normalfont
Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of f-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also aim at optimizing the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}

>>>>

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. Often, the optimal implementation strategy for these operations is to be determined out of two alternatives. To illustrate the concept, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. The tensor operation will reduce to $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on loop $L_i$. To minimize the operation count for sub-expressions of this kind, we have two options.
\begin{Strategy}
\label{strategy:i}
Eliminating temporal sharing through generalized code motion.
\end{Strategy}
\begin{Strategy}
\label{strategy:ii}
Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
\end{Strategy}
In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count, since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed. On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$. In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion. Note that a second example of application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}. These examples motivate the introduction of a particular class of expressions.
\begin{Def}[Structured expression]
\label{def:struct-expr}
An expression is \textit{structured} with respect to a set of loops $\Lambda$ if and only if there is no ambiguity in the elimination of spatial sharing for the symbols depending on $\Lambda$.
\end{Def}
This leads us to the following, fundamental observation.
\begin{Prop}
\label{prop:multi-struct}
In a multilinear loop nest $\Lambda$, all of the expressions are structured with respect to $\Lambda$.
\end{Prop}
In finite element integration loop nests, for example, any factorization of test and trial functions will always result in exposing temporal sharing (Strategy~\ref{strategy:ii}). Conversely, in a generic loop nest $\Lambda$, the instances of a symbol depending on $\Lambda$ may appear everywhere (e.g., twice in a summand and only once in all other summands; as argument of a higher level function; in the denominator of a division), thus posing the challenge of finding the optimal factorization that maximizes temporal sharing. Proposition~\ref{prop:multi-struct} played a key role in the development of Algorithm~\ref{algo:sharing-elimination}. Finally, we observe that structured sub-expressions sometimes characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. It makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\end{itemize}
Examples will be provided in Section~\ref{sec:se-examples}.

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree to group identical terms within the same sub-expression. \\ 
\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.

@@

\item For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}. \\
\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied. This is a consequence of the expression being structured with respect to $[L_j, L_k]$.}

@@

\begin{Rem*}
\normalfont
Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of f-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also attempt to optimize the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}
